The Evolution of Phenotypically Invariant Genetic Networks

### 2017-02-28 kalman_walk.tex
removed redundant equations.
added bio examples, including GRN to Metabolic Network systems.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

@@

Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).

  \begin{align*}
    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
  \end{align*}
For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
  \begin{align*}
    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
  \end{align*}

 The transfer function at $z=0$ for any GRN with mutations is, 
  \begin{align*}
    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
  \end{align*}

@@

\section{Notes on what to do when allowing deleterious mutations}

>>>>

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

@@

%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
%
%  \begin{align*}
%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
%  \end{align*}
%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
%  \begin{align*}
%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
%  \end{align*}
%
 The transfer function at $z=0$ for any GRN with mutations is, 
  \begin{align*}
    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
  \end{align*}

@@

\section{Biological Examples}

  \begin{example}
    Negative feedback and Positive feedback systems:
    \begin{align*}
      T(z) = \frac{H(z)}{1 + G(z)H(z)} \\
      T(z) = \frac{H(z)}{1 - G(z)H(z)} \\
    \end{align*}
  \end{example}

  \begin{example}
    Systems in serial ($u \rightarrow \Sigma_{1} \{A_{1},B_{1},C_{1} \} \rightarrow \Sigma_{2} \{A_{2},B_{2},C_{2} \} \rightarrow y$):
    \begin{align*}
      \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} &= \left[ \begin{array}{c|c}
        A_{1} & 0 \\
          \hline
          B_{2}C_{2} & A_{2} \\
        \end{array} \right] \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix}  + \begin{bmatrix} B_{1} \\ 0 \end{bmatrix} u \\
          y &= \begin{bmatrix} 0 & C_{2} \end{bmatrix} \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} .
    \end{align*}

  Systems in serial could be applied to model a gene network $A_{1}$ and the subsequent degradation rates of its outputs at rate $A_{2}$.
  \end{example}

  \begin{example}
  \begin{align*}
    \dot{x} &= \begin{bmatrix} + & - \\ 0 & + \end{bmatrix} \vec{x} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{u} \\
      y(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{x} \\
        E(t) &= \begin{bmatrix} -y_{1}(t) & 0 & 0 \\ y_{1}(t) & -y_{2}(t) & 0 \\ -\sigma_{1} y_{1}(t) & (1 - \sigma_{2})y_{2}(t) & 1 \end{bmatrix} \vec{\zeta} + \begin{bmatrix} u_{1} \\ u_{2} \\ \epsilon \end{bmatrix} .
  \end{align*}
    Where $u_{1}$ is input of \emph{lactose}, $u_{2}$ is the input of \emph{glucose}, $y_{1}$ is the enzyme \emph{lactase} and $y_{2}$ is enzyme \emph{``glucase''}. $E(t)$ is metabolism of the organism, converting sugars to energy as a function of enzyme concentrations. Lactase converts lactose to glucose, and the enzyme is produced at energy cost $\sigma_{1}$; glucase converts glucose to energy and is produced at cost $\sigma_{2}$. Fitness is a function of the total energy of the system, given inputs, possibly: $\int_{0}^{T}E(t)_{3} dt$ after some time $T$ has passed. Or perhaps, once the total metabolic energy produced passes some threshold, the organism undergoes fission. Thus, fitness selects for maximizing energy as quickly as possible. 
  \end{example}

  \begin{example} 
    Gene Regulatory and Metabolic Network Cascade:

    Gene Regulatory Network
    \begin{align*}
      \dot{x}(t) &= A x(t) + B_{1} u(t) \\
      y(t) &= C_{1} x(t)
    \end{align*}
      Metabolism
      \begin{align*}
      \dot{\zeta}(t) &= M\big(y(t), \sigma \big) \zeta(t) + B_{2} \begin{bmatrix} u(t) \\ v(t) \end{bmatrix} \\
        m(t) &= C_{2} \zeta(t) \\
      b(t) &= C_{3} \zeta(t)
      \end{align*}
      Fitness
      \begin{align*}
      \Phi(m(t), p(t))
    \end{align*}

    The GRN produces time varying proteins involved in the metabolism. The metabilic network (MN), is a function of protein concentrations $y(t)$ and production costs $\sigma$. The metabolic network produces energy $m(t)$ and toxic byproducts $b(t)$. Fitness $\Phi$, or maybe something more concrete, like fission time and/or cell death, is a function of both energy $m(t)$ and toxins $b(t)$. 
  \end{example}
\section{Notes on what to do when allowing deleterious mutations}
